---
title: "Estimating CGM Mean First-Passage Times"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating CGM Mean First-Passage Times}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`CGMmfpt` estimates the time required for a three-state glucose process to
first return to its middle state. The empirical method uses hard state
assignments. The HMM uses beta-uniform mixture emissions with overlapping
supports around the two state boundaries.

## Synthetic data

The package's example generator produces synthetic data and does not reproduce
the study dataset or manuscript results.

```{r data}
library(CGMmfpt)

cgm <- simulate_cgm_data(
  n_participants = 4,
  n_per_period = 72,
  seed = 11
)
head(cgm)
```

## Pooled emission templates

Emission distributions are estimated from a reference sample and then held
fixed in participant-level HMMs. Here, the baseline observations are pooled
within age group.

```{r templates}
baseline <- cgm[cgm$period == "baseline", ]
cutoffs <- list(adults = c(70, 180), children = c(70, 180))

templates <- fit_emission_templates(
  baseline,
  glucose = "glucose",
  stratum = "age_group",
  cutoffs = cutoffs,
  overlap = 10,
  min_state_n = 5
)
summary(templates)
```

The `overlap` argument is the half-width in mg/dL on either side of a cutoff.
Within each overlap, normalized baseline prevalences weight the two adjacent
emission scores.

## HMM estimates

```{r hmm}
hmm_results <- estimate_hmm_mfpt(
  cgm,
  templates,
  id = "participant_id",
  time = "timestamp",
  glucose = "glucose",
  group = c("period", "treatment"),
  stratum = "age_group",
  min_obs = 40,
  max_iter = 30
)
hmm_results
```

The HMM is fitted separately to each participant-period series. Gaps longer
than six minutes begin new segments by default. Emission parameters remain
fixed; the segment-start probabilities and transition matrix are estimated by
Baum-Welch.

## Empirical estimates

```{r empirical}
empirical_results <- estimate_empirical_mfpt(
  cgm,
  id = "participant_id",
  time = "timestamp",
  glucose = "glucose",
  group = c("period", "treatment"),
  stratum = "age_group",
  cutoffs = cutoffs,
  reference_by = c("age_group", "period"),
  prior_strength = 20,
  min_obs = 40
)
empirical_results
```

The empirical transition counts can be used without smoothing by setting
`prior_strength = 0`. With smoothing, each count row receives pseudocounts
proportional to the pooled transition probabilities in its reference stratum.

## One fitted series

```{r one-fit}
one <- cgm[cgm$participant_id == "P001" & cgm$period == "rct", ]
fit <- fit_hmm_cgm(
  one$glucose,
  one$timestamp,
  templates,
  stratum = one$age_group[1],
  return_posterior = TRUE
)
fit$transition_matrix
fit$mfpt[c("lower", "upper")]
head(fit$posterior)
```

MFPT values are reported in minutes. Infinite values indicate that the fitted
chain does not reach the middle state with probability one.
